---
title: "Optimising Sequencing Budgets for Pathogen Surveillance"
subtitle: "Design-adjusted inference and Neyman allocation under unequal sampling"
author: "Cuiwei Gao"
date: "2026-04-19"
categories: [genomic-surveillance, survey-sampling, public-health]
knitr:
  opts_chunk:
    message: false
    warning: false
    dev: "png"
    fig.align: "center"
    dpi: 150
---

## Motivation

Pathogen surveillance systems sequence unevenly. A European programme
might have Denmark sequencing 12 % of confirmed cases while Romania
sequences 0.3 % — a **24×** difference. If you estimate a lineage's
continental prevalence by *pooling* the sequenced samples and counting,
your answer is dominated by Denmark's local epidemic, regardless of
what is actually circulating in the EU.

The same asymmetry appears in *time*: a fast-growing variant that
sequences are still travelling through the lab looks rarer than it
really is by the time it's reported. And the *budget* question —
"we have $N$ sequences a week; how should we split them across
regions to best detect an emerging lineage?" — has a formal answer
from survey-sampling theory (Neyman 1934), not an informal one.

This case study applies **[`survinger`](https://CRAN.R-project.org/package=survinger)**
(CRAN 0.1.1) to a realistic simulated surveillance dataset shipped
with the package, and walks through four decisions every
surveillance programme implicitly makes:

1.  How to **estimate prevalence** when sequencing rates differ.
2.  How to **nowcast** around report delays.
3.  How to **allocate** a fixed sequencing budget across regions.
4.  How to **size** a programme for reliable early detection.

## Data

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(survinger)
  library(ggplot2)
  library(dplyr)
})

data(sarscov2_surveillance)
ss <- sarscov2_surveillance
```

`sarscov2_surveillance` is a simulated 6-month programme across 5
regions with deliberately *unequal* sequencing rates — the package
ships the ground truth alongside the observed data so the effect of
methodological choices can be measured, not guessed.

```{r}
#| label: tbl-rates
ss$population |>
  transmute(Region = region,
            `Positive cases` = n_positive,
            `Sequenced` = n_sequenced,
            `Rate` = scales::percent(seq_rate, accuracy = 0.1),
            `Population` = scales::comma(pop_total))
```

The sequencing rate spans **`r scales::percent(min(ss$population$seq_rate), accuracy = 0.1)`**
to **`r scales::percent(max(ss$population$seq_rate), accuracy = 0.1)`**
across the five regions — a roughly **`r round(max(ss$population$seq_rate) / min(ss$population$seq_rate))`×**
gap.

## Step 1 — A surveillance design object

```{r}
#| label: design
d <- surv_design(
  data            = ss$sequences,
  strata          = ~ region,
  sequencing_rate = ss$population[, c("region", "seq_rate")],
  population      = ss$population
)
d
```

The design object carries the sampling weights (inverse of sequencing
rates). Every estimator downstream uses them.

## Step 2 — Naive vs design-adjusted prevalence

For an emerging lineage such as **BA.2.86**, how much does
sequencing-rate inequality bias a naive estimate? The package ships
the truth (simulation ground truth), so we can plot naive versus
Hajek-adjusted estimates against it directly.

```{r}
#| label: prev-compare
naive <- surv_naive_prevalence(d, lineage = "BA.2.86")$estimates |>
  mutate(method = "Naive (unweighted)")

hajek <- surv_lineage_prevalence(d, lineage = "BA.2.86",
                                 method = "hajek")$estimates |>
  mutate(method = "Design-adjusted (Hajek)")

truth <- ss$truth |>
  filter(lineage == "BA.2.86") |>
  group_by(epiweek) |>
  summarise(
    prevalence = weighted.mean(true_prevalence,
                               w = ss$population$pop_total[match(region, ss$population$region)]),
    .groups    = "drop"
  ) |>
  rename(time = epiweek) |>
  mutate(method = "Truth (population-weighted)")

compare <- bind_rows(
  naive |> select(time, prevalence, ci_lower, ci_upper, method),
  hajek |> select(time, prevalence, ci_lower, ci_upper, method),
  truth |> mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(time = as.Date(paste0(sub("-W", "-", time), "-1"), format = "%Y-%U-%u"))
```

```{r}
#| label: fig-bias
#| fig-cap: "Estimated BA.2.86 prevalence over 26 epiweeks: naive (unweighted) versus design-adjusted (Hajek) versus population-weighted ground truth. The naive estimator is dominated by the high-sequencing regions (A, B) and systematically misrepresents the population trajectory. Hajek tracks the truth within the 95 % CI. Ribbons are the 95 % CIs."
#| fig-height: 4.5
ggplot(compare, aes(x = time, y = prevalence, colour = method, fill = method)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
              alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c(
    "Naive (unweighted)"          = "#C62828",
    "Design-adjusted (Hajek)"     = "#1565C0",
    "Truth (population-weighted)" = "#2E7D32"
  ), name = NULL) +
  scale_fill_manual(values = c(
    "Naive (unweighted)"          = "#C62828",
    "Design-adjusted (Hajek)"     = "#1565C0",
    "Truth (population-weighted)" = "#2E7D32"
  ), name = NULL) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "BA.2.86 prevalence") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "top")
```

```{r}
#| label: bias-summary
#| echo: false
naive_vals <- naive$prevalence
hajek_vals <- hajek$prevalence
truth_vals <- truth$prevalence[match(naive$time, truth$time)]
mae_naive <- mean(abs(naive_vals - truth_vals), na.rm = TRUE)
mae_hajek <- mean(abs(hajek_vals - truth_vals), na.rm = TRUE)
```

Averaged across the 26 weeks, the naive estimator is off by
**`r scales::percent(mae_naive, accuracy = 0.1)`** (mean absolute
error versus the known population truth); the Hajek estimator is
off by **`r scales::percent(mae_hajek, accuracy = 0.1)`** —
a roughly **`r round(mae_naive / mae_hajek, 1)`×** reduction.
The entire bias is a *design* issue; no model tuning is involved.

## Step 3 — Delay-adjusted nowcasting

Sequences arrive late. The `report_date` comes days to weeks after
`collection_date`, so the most-recent weeks in a "raw" estimate are
always under-counted. `surv_estimate_delay()` fits a delay
distribution (negative binomial here, more flexible than Poisson
because delay tails are heavy) and `surv_nowcast_lineage()` corrects
for it.

```{r}
#| label: nowcast
#| results: "hide"
dfit    <- surv_estimate_delay(d, distribution = "negbin")
adjusted <- surv_adjusted_prevalence(d, delay_fit = dfit,
                                     lineage = "BA.2.86")
```

```{r}
#| label: fig-nowcast
#| fig-cap: "Raw Hajek prevalence (blue) versus delay-adjusted prevalence (red). The delay adjustment uses a fitted negative-binomial report-delay distribution to up-weight the most-recent weeks whose reports are still arriving; without the correction, a rising variant looks flat at the right-hand edge. Ribbon is the 95 % CI of the delay-adjusted estimate."
#| fig-height: 4
parse_epiweek <- function(x) {
  as.Date(paste0(sub("-W", "-", x), "-1"), format = "%Y-%U-%u")
}

adj_df <- adjusted$estimates |>
  mutate(time = parse_epiweek(time))
raw_df <- hajek |>
  mutate(time = parse_epiweek(time))

ggplot() +
  geom_ribbon(data = adj_df,
              aes(x = time, ymin = ci_lower, ymax = ci_upper),
              fill = "#C62828", alpha = 0.18) +
  geom_line(data = raw_df, aes(x = time, y = prevalence),
            colour = "#1565C0", linewidth = 0.9) +
  geom_line(data = adj_df, aes(x = time, y = prevalence),
            colour = "#C62828", linewidth = 0.9, linetype = "dashed") +
  annotate("text", x = min(raw_df$time) + 40,
           y = max(raw_df$prevalence, na.rm = TRUE) * 0.9,
           label = "Hajek (raw)", colour = "#1565C0",
           hjust = 0, size = 3.8, fontface = "bold") +
  annotate("text", x = max(adj_df$time),
           y = max(adj_df$prevalence, na.rm = TRUE),
           label = "Delay-adjusted",
           colour = "#C62828", hjust = 1, vjust = 1.4,
           size = 3.8, fontface = "bold") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "BA.2.86 prevalence") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## Step 4 — Optimal allocation of a fixed budget

Suppose the programme is capped at 500 sequences per week. How should
those 500 be split across the five regions to *minimise the MSE of
overall prevalence*? `surv_optimize_allocation()` returns the Neyman
allocation that balances within-stratum variance and population size.

```{r}
#| label: alloc
alloc <- surv_optimize_allocation(d, objective = "min_mse",
                                  total_capacity = 500L)
alloc$allocation
```

```{r}
#| label: fig-alloc
#| fig-cap: "Current sequencing allocation (observed counts normalised to 500) versus Neyman-optimal allocation for minimising MSE of overall prevalence. Region D is under-sequenced relative to its information content; Regions A and E are near-optimal."
#| fig-height: 4
current <- ss$population |>
  mutate(proportion = n_sequenced / sum(n_sequenced),
         n_allocated = round(proportion * 500)) |>
  transmute(region, n_allocated, proportion, allocation = "Current (scaled to 500)")

optimal <- alloc$allocation |>
  mutate(allocation = "Neyman optimal (min-MSE)")

alloc_df <- bind_rows(current, optimal)

ggplot(alloc_df, aes(x = region, y = proportion, fill = allocation)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = scales::percent(proportion, accuracy = 1)),
            position = position_dodge(width = 0.7),
            vjust = -0.3, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("Current (scaled to 500)" = "#546E7A",
                               "Neyman optimal (min-MSE)" = "#1565C0"),
                    name = NULL) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL, y = "Share of 500 weekly sequences") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "top")
```

## Step 5 — How large a programme detects a rare lineage?

Finally: at current capacity, what is the probability of catching a
hypothetical lineage at 1 % true prevalence within a single week?
`surv_detection_probability()` answers this per-stratum under the
existing allocation.

```{r}
#| label: detection
det <- surv_detection_probability(d, true_prevalence = 0.01,
                                  n_periods = 1L,
                                  detection_threshold = 1L)
det$by_stratum
```

```{r}
#| label: detection-curve
#| results: "hide"
prev_grid <- seq(0.002, 0.05, length.out = 20)
curve_df <- do.call(rbind, lapply(prev_grid, function(p) {
  r <- surv_detection_probability(d, true_prevalence = p,
                                  n_periods = 1L,
                                  detection_threshold = 1L)
  data.frame(true_prevalence = p, p_detect = r$overall)
}))
```

```{r}
#| label: fig-detection
#| fig-cap: "Overall detection probability in a single week as a function of a lineage's true prevalence, at current sequencing capacity and allocation. A variant at 1 % prevalence has roughly 40 % probability of being observed in any given week; to cross 90 % detection probability requires true prevalence above ~3 %."
#| fig-height: 4
ggplot(curve_df, aes(x = true_prevalence, y = p_detect)) +
  geom_line(colour = "#1565C0", linewidth = 1) +
  geom_hline(yintercept = 0.9, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  labs(x = "True prevalence of a target lineage",
       y = "Probability of ≥1 observation in a week") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## Interpretation

**Bias is a design problem, not a model problem.** Switching from
naive pooling to the Hajek estimator cut MAE by roughly
`r round(mae_naive / mae_hajek, 1)`× on this dataset; the cost was
one line of code and the sequencing-rate table that the programme
already maintains for its own reporting. Any downstream use of
surveillance estimates — nowcasts, forecasts, growth-advantage
fits — should start from a design-adjusted estimator, not a raw
pooled one.

**Budgets aren't obvious.** The current allocation was proportional
to sequencing *rates* (which track programme capacity). The
Neyman-optimal allocation redistributes toward the stratum whose
variance contribution to the overall estimate is largest. For this
programme, that is Region D — currently under-sequenced relative to
its information content. Reallocating weekly capacity accordingly
would be a zero-cost improvement to overall-prevalence precision.

**Early-warning realism.** At current capacity, detecting a 1 %-prevalence
emerging variant within a single week is essentially a coin-flip.
Reaching a 90 % one-week detection probability requires either more
sequences per week or for the variant to already be past ~3 %
prevalence. That framing — a *detection probability*, not a
*detection guarantee* — is the one a programme should be able to
produce from its design alone, and `survinger` makes it one call.

## About this analysis

::: {.callout-note appearance="minimal" collapse="true"}
## Environment
- **Author:** Cuiwei Gao
- **Date:** 2026-04-19
- **Package:** `survinger` v`r packageVersion("survinger")`
- **Data:** `sarscov2_surveillance` (simulated 5-region, 26-week SARS-CoV-2 programme shipped with `survinger`)
- **Source:** [github.com/CuiweiG/portfolio](https://github.com/CuiweiG/portfolio/tree/main/03-surveillance-design)

```{r}
#| label: session
sessionInfo()
```
:::
